Constraints on Cosmic Rays, Magnetic Fields, and Dark Matter from Gamma-Ray Observations of the Coma Cluster of Galaxies with VERITAS and \Fermi

### 2012-02-27 ms.tex
Martin's edits before submission

\linenumbers

@@

%\title{VHE Gamma-Ray Observations of the Coma Cluster of Galaxies with VERITAS}
%\title{Constraints on Cosmic-Ray Particle Populations, Magnetic Fields, and Dark Matter from VHE Gamma-Ray Observations of the Coma Cluster of Galaxies with VERITAS}
\title{Constraints on Cosmic Rays, Magnetic Fields, and Dark Matter from Gamma-Ray Observations of the Coma Cluster of Galaxies with VERITAS and \Fermi}
\author{
T. Arlen\altaffilmark{1},
M. Beilicke\altaffilmark{2},
W. Benbow\altaffilmark{3},
V. Bugaev\altaffilmark{2},
K. Byrum\altaffilmark{4},
A. Cannon\altaffilmark{5},
A. Cesarini\altaffilmark{6},
L. Ciupik\altaffilmark{7},
E. Collins-Hughes\altaffilmark{5},
W. Cui\altaffilmark{8},
R. Dickherber\altaffilmark{2},
J. Dumm\altaffilmark{9},
Q. Feng\altaffilmark{8},
G. Finnegan\altaffilmark{10},
L. Fortson\altaffilmark{9},
A. Furniss\altaffilmark{11},
N. Galante\altaffilmark{3},
J. Grube\altaffilmark{7},
G. Gyuk\altaffilmark{7},
H. Huan\altaffilmark{12},
G. Hughes\altaffilmark{13},
T. B. Humensky\altaffilmark{14},
P. Kaaret\altaffilmark{15},
N. Karlsson\altaffilmark{9,27},
Y. Khassen\altaffilmark{5},
D. Kieda\altaffilmark{10},
H. Krawczynski\altaffilmark{2},
G. Maier\altaffilmark{13},
P. Majumdar\altaffilmark{1},
S. McArthur\altaffilmark{2},
A. McCann\altaffilmark{16},
P. Moriarty\altaffilmark{17},
M. Orr\altaffilmark{18},
A. N. Otte\altaffilmark{19},
J. S. Perkins\altaffilmark{20,21},
M. Pohl\altaffilmark{22,13},
H. Prokoph\altaffilmark{13},
P. T. Reynolds\altaffilmark{23},
E. Roache\altaffilmark{3},
J. Ruppel\altaffilmark{22,13},
D. B. Saxon\altaffilmark{24},
G. H. Sembroski\altaffilmark{8},
C. Skole\altaffilmark{13},
A. W. Smith\altaffilmark{10},
I. Telezhinsky\altaffilmark{22,13},
G. Te\v{s}i\'{c}\altaffilmark{16},
S. Thibadeau\altaffilmark{2},
M. Vivier\altaffilmark{24},
J. E. Ward\altaffilmark{5},
R. Welsing\altaffilmark{13}
}

\and
\author{S. Federici\altaffilmark{13,22}, C. Pfrommer\altaffilmark{25,27},
A. Pinzke\altaffilmark{26}}

\altaffiltext{1}{Department of Physics and Astronomy, University of California, Los Angeles, CA 90095, USA}
\altaffiltext{2}{Department of Physics, Washington University, St. Louis, MO 63130, USA}
\altaffiltext{3}{Fred Lawrence Whipple Observatory, Harvard-Smithsonian Center for Astrophysics, Amado, AZ 85645, USA}
\altaffiltext{4}{Argonne National Laboratory, 9700 S. Cass Avenue, Argonne, IL 60439, USA}
\altaffiltext{5}{School of Physics, University College Dublin, Belfield, Dublin 4, Ireland}
\altaffiltext{6}{School of Physics, National University of Ireland Galway, University Road, Galway, Ireland}
\altaffiltext{7}{Astronomy Department, Adler Planetarium and Astronomy Museum, Chicago, IL 60605, USA}
\altaffiltext{8}{Department of Physics, Purdue University, West Lafayette, IN 47907, USA }
\altaffiltext{9}{School of Physics and Astronomy, University of Minnesota, Minneapolis, MN 55455, USA}
\altaffiltext{10}{Department of Physics and Astronomy, University of Utah, Salt Lake City, UT 84112, USA}
\altaffiltext{11}{Santa Cruz Institute for Particle Physics and Department of Physics, University of California, Santa Cruz, CA 95064, USA}
\altaffiltext{12}{Enrico Fermi Institute, University of Chicago, Chicago, IL 60637, USA}
\altaffiltext{13}{DESY, Platanenallee 6, 15738 Zeuthen, Germany}
\altaffiltext{14}{Physics Department, Columbia University, New York, NY 10027, USA}
\altaffiltext{15}{Department of Physics and Astronomy, University of Iowa, Van Allen Hall, Iowa City, IA 52242, USA}
\altaffiltext{16}{Physics Department, McGill University, Montreal, QC H3A 2T8, Canada}
\altaffiltext{17}{Department of Life and Physical Sciences, Galway-Mayo Institute of Technology, Dublin Road, Galway, Ireland}
\altaffiltext{18}{Department of Physics and Astronomy, Iowa State University, Ames, IA 50011, USA}
\altaffiltext{19}{School of Physics \& Center for Relativistic Astrophysics, Georgia Institute of Technology, 837 State Street NW, Atlanta, GA 30332-0430}
\altaffiltext{20}{CRESST and Astroparticle Physics Laboratory NASA/GSFC, Greenbelt, MD 20771, USA.}
\altaffiltext{21}{University of Maryland, Baltimore County, 1000 Hilltop Circle, Baltimore, MD 21250, USA.}
\altaffiltext{22}{Institut f\"ur Physik und Astronomie, Universit\"at Potsdam, 14476 Potsdam-Golm,Germany}
\altaffiltext{23}{Department of Applied Physics and Instrumentation, Cork Institute of Technology, Bishopstown, Cork, Ireland}
\altaffiltext{24}{Department of Physics and Astronomy and the Bartol Research Institute, University of Delaware, Newark, DE 19716, USA}

% Christoph, Anders and Simone:
\altaffiltext{25}{Heidelberg Institute for Theoretical Studies, Schloss-Wolfsbrunnenweg 35, D-69118 Heidelberg, Germany}
\altaffiltext{26}{Department of Physics, University of California, Santa Barbara, CA 93106, USA}
\altaffiltext{27}{Corresponding authors: N. Karlsson, nkarlsson@physics.umn.edu \& C. Pfrommer, christoph.pfrommer@h-its.org}

@@

\begin{abstract}
  \linenumbers Observations of radio halos and relics in galaxy clusters indicate efficient electron
  acceleration processes; the same processes are even more efficient for protons. Since protons have
  a cooling time which exceeds the Hubble time in clusters, they can accumulate suggesting that
  clusters may also be sources of very high-energy (VHE; $E>100$ GeV) gamma-ray emission resulting
  from the interactions of highly relativistic cosmic-ray (CR) hadrons with the intracluster gas. We
  report here on VHE gamma-ray observations of the Coma galaxy cluster with the VERITAS array of
  imaging Cherenkov telescopes, with complementing \Fermi Gamma-ray Space Telescope observations at
  GeV energies. No significant gamma-ray emission from the Coma cluster was detected. Integral flux
  upper limits at the 99\% confidence level were measured to be on the order of $2-5\times 10^{-8}$
  ph.  m$^{-2}$ s$^{-1}$ (VERITAS, $>$220 GeV) and $\sim 3\times 10^{-6}$ ph. m$^{-2}$ s$^{-1}$
  (\Fermi, 1-3 GeV), respectively. For VERITAS (\Fermi), the limits are a factor of 60 (20) higher
  than model-predicted minimum gamma-ray fluxes, which assume the radio-emitting electrons of the
  Coma radio halo are all secondaries from cosmic-ray interactions. We use the gamma-ray flux upper
  limits to constrain the populations of CRs and magnetic fields in Coma. Using an analytical
  approach motivated by cosmological cluster simulations, the CR-to-thermal pressure ratio is
  constrained to be $< 16\%$ from VERITAS data and $< 2.5\%$ from \Fermi data (averaged within the
  virial radius). Assuming that the radio-emitting electrons of the Coma halo result from hadronic
  CR interactions, the observations imply a lower limit on the central magnetic field in Coma of
  $1$--$3\mu$G, depending on how the magnetic field declines with radius and on the gamma-ray
  spectral index. Since these values are well below those inferred by Faraday rotation measurements
  in Coma, this makes the hadronic model a very plausible explanation of the Coma radio
  halo. Finally, since galaxy clusters are dark matter (DM) dominated, the VERITAS upper limits have
  been used to place constraints on the thermally averaged product of the total self-annihilation
  cross section and the relative velocity of the DM particles, $\expval{\sigma v}$. This is found to
  be about $10^{-20} - 10^{-21}$ cm$^{3}$ s$^{-1}$ for a DM particle mass $m_{\chi} = 300$
  GeV/c$^{2}$.
\end{abstract}

@@

The ICM is a hot ($T\sim 10^{8}$ K) plasma emitting thermal bremsstrahlung in the soft X-ray regime
\citep[see, e.g.,][]{article:Petrosian:2001}. This plasma has been heated primarily through
collisionless structure-formation shocks that form as a result of the hierarchical merging and
accretion processes. Such shocks and turbulence in the ICM gas in combination with intra-cluster
magnetic fields also provide a means to accelerate particles efficiently \citep[see,
e.g.,][]{article:ColafrancescoBlasi:1998, article:Ryu_etal:2003}. Many clusters feature megaparsec
scale halos of nonthermal radio emission, indicative of a population of relativistic electrons and
positrons\footnote{From here on electrons and positrons are collectively referred to as electrons.}
and magnetic fields permeating the ICM \citep{article:Cassano_etal:2010}. There are two competing
theories to explain radio halos. In the ``hadronic model'', the radio-emitting
electrons are produced in inelastic collisions of cosmic-ray (CR) ions with the thermal gas of the
ICM \citep{article:Dennison:1980, article:EnsslinPfrommerMiniatiSubramanian:2011}. In the
``re-acceleration model'', a long-lived pool of 100-MeV electrons---previously accelerated by
formation shocks, galactic winds, or jets of active galactic nuclei (AGN)---interacts with plasma
waves that are excited during states of strong ICM turbulence, e.g., after a cluster merger. This
may result in second order Fermi acceleration and may produce energetic electrons ($\sim 10$ GeV)
sufficient to explain the observable radio emission \citep{article:SchlickeiserSieversThiemann:1987,
article:BrunettiLazarian:2010}. Observations of possibly nonthermal emission from clusters
in the extreme ultraviolet \citep[EUV; ][]{article:SarazinLieu:1998} and hard X-rays
\citep{article:RephaeliGruber:2002, article:Fusco-Femiano_etal:2004, article:Eckert_etal:2007} may
provide further indication of relativistic particle populations in clusters, although the
interpretation of these observations as nonthermal diffuse emission has been disputed on the basis
of more sensitive observations \citep[see, e.g.,][]{article:Ajello_etal:2009,
article:Ajello_etal:2010, article:Wik_etal:2009}.

@@

Gamma-ray emission can also be produced by Compton up-scattering of ambient photons, for example
CMB photons, on ultra-relativistic electrons. Those electrons can either be secondaries from the
above mentioned CR interactions or injected into the ICM by powerful cluster members and
further accelerated by diffusive shock acceleration or turbulent reacceleration processes
\citep[][and references therein]{article:SchlickeiserSieversThiemann:1987}.

@@

\section{VERITAS Observations and Analysis}
The VERITAS gamma-ray detector \citep{article:Weekes_etal:2002} is an array of four 12 m-diameter
imaging atmospheric Cherenkov telescopes \citep{article:Holder_etal:2006} located at an
altitude of $\sim$1250 m a.s.l. at the Fred Lawrence Whipple Observatory in southern Arizona
(31$^{\circ}$~40\arcmin~30\arcsec~N, 110$^{\circ}$~57\arcmin~07\arcsec~W). Each of the telescopes
is equipped with a 499-pixel camera covering a 3.5$^{\circ}$ field of view. The array, completed in
the fall of 2007, is designed to detect gamma-ray emission from astrophysical objects in the
energy range from 100 GeV to more than 30 TeV. The energy resolution is $\sim 15$\% and the angular
resolution (68\% containment) is $\sim 0.1^{\circ}$ per event at 1 TeV. At the time of the Coma
cluster observations, the sensitivity of the array allowed for detection of a point source with a
flux of 1\% of the steady Crab Nebula flux above 300 GeV at the confidence level of five standard
deviations ($5\sigma$) in under 45 hours.\footnote{The integral flux sensitivity above 300 GeV was
improved by about 30\% with the relocation of one telescope in the summer of 2009.}

@@

@@

\subsection{VERITAS Results}
No significant excess of VHE gamma rays from the Coma cluster was detected with VERITAS, as
illustrated by the $\theta^{2}$ distribution shown in Figure \ref{fig:thetasq}. The $\theta^{2}$
distribution is a radial comparison of the ON-source event distribution with the estimated
background. It extends out to 0.42 square degrees to cover both the case of point-like and extended
emission from the core of the Coma cluster. The $\theta^{2}$ distributions for the member galaxies
also considered in this work are very similar to that in Figure \ref{fig:thetasq} and show no
excess of gamma rays. A 99\% confidence level upper limit is calculated for each region of
interest using the method described by \citet{article:Rolke_etal:2005} assuming a
Gaussian-distributed background. A lower bound of zero is imposed on the signal rate to handle the
negative excesses seen for several of the regions of interest. Figure \ref{fig:sigdist} shows the
distribution of significances over the VERITAS skymap, which is well fit by a Gaussian with a mean
close to zero and a standard deviation within a few percent of unity.

Table \ref{table:results} lists the upper limits for the selected regions of interest shown in
Table \ref{table:roi}. These upper-limit calculations depend on the gamma-ray spectrum, which in
this work is assumed to be a power law in energy, $F\propto E^{-\alpha}$, where the spectral index
$\alpha$ was allowed to have a value of 2.1, 2.3, or 2.5.

\section{\Fermi-LAT Analysis and Results}
The Large Area Telescope (LAT) on board the Fermi Gamma-ray Space Telescope (Fermi) has observed
the Coma cluster in all-sky survey mode since its launch in June 2008. The Fermi-LAT is sensitive
to gamma rays in the 20 MeV to $\sim300$ GeV energy range and is very complementary to the VERITAS
observations. Ackermann et al. (2010) reported on the search for gamma-ray emission from
thirty-three galaxy clusters in the data from the first 18 months, including the Coma cluster, for
which an upper limit of $4.58\times10^{-9}$ ph. cm$^{-2}$ s$^{-1}$ in the 0.2 to 100 GeV energy band was reported. This limit is expected to improve as the exposure is increased. In this work an
updated analysis which includes additional data taken during the period 2010 February 5 to 2011 
June 10 is presented as a complement to the VERITAS results.

@@

The remaining background includes twenty-eight sources listed in the 11 months catalog of Fermi-LAT
data (Abdo et al. 2010), and the LAT standard Galactic diffuse emission component
\verb+gll_iem_v02.fit+ along with the corresponding isotropic template
\verb+isotropic_iem_v02.txt+ that accounts for extragalactic emission and residual instrumental
background. All of the twenty-eight sources are modeled as point sources with power-law energy
spectra. Fourteen of them lie outside the ROI and their spatial and spectral parameters were being
kept fixed at the values given in the catalog. The parameters of the remaining sources, being
inside the ROI, and the normalization of the diffuse components were left to freely vary. The
gamma-ray emission from the Coma cluster is assumed to follow a power-law distribution with a
spectral index $\alpha$, which is allowed to have a value of 2.1, 2.3, or 2.5.

No significant gamma-ray signal was detected. Values of the significance greater than one were only
obtained for the higher energy bin, where the test statistic (TS) is $\sim2$ for the point source
model, $\sim4$ for the disk model with r$=0.2^{\circ}$ and $\sim6$ for the disk model with
r$=0.4^{\circ}$.

@@

\section{Gamma Ray Emission from Cosmic Rays}
We decided to adopt a multifaceted approach to constrain the CR-to-thermal pressure distribution in
the Coma cluster using the upper limits derived from the VERITAS and \Fermi-LAT data in this work.
This approach includes (1) a simplified multi-frequency analytical model that assumes a constant
CR-to-thermal energy density and a power-law spectrum in momentum, (2) an analytic model derived
from cosmological hydrodynamical simulations of the formation of galaxy clusters, and (3) an
model that uses the observed luminosity of the giant radio halo in Coma to place a lower limit on
the expected gamma-ray flux in the hadronic model -- where the radio-emitting electrons are
secondaries from CR interactions and which is independent of the magnetic field distribution.  This
last approach translates into a minimum CR pressure which, if challenged by tight gamma-ray
limits/detections, permits scrutiny of the hadronic interaction model of the formation of giant
radio halos. Alternatively, realizing a spatial CR distribution that is consistent with the flux
upper limits, and requiring the model to match the observed radio data, enables us to derive a
lower limit on the magnetic field distribution. We stress again that this approach assumes the
validity of the hadronic interaction model. Modeling the CR distribution through different
techniques enables us to bracket our lack of understanding about the underlying plasma physics that
shapes the CR distribution hence to reflect the Bayesian priors that are imposed on the modeling
\citep[see][for a discussion]{article:PinzkePfrommerBergstrom}.

>>>>

%\linenumbers

@@

%\title{VHE Gamma-Ray Observations of the Coma Cluster of Galaxies with VERITAS}
%\title{Constraints on Cosmic-Ray Particle Populations, Magnetic Fields, and Dark Matter from VHE Gamma-Ray Observations of the Coma Cluster of Galaxies with VERITAS}
\title{Constraints on Cosmic Rays, Magnetic Fields, and Dark Matter from Gamma-Ray Observations of the Coma Cluster of Galaxies with VERITAS and \Fermi}
\author{
T.~Arlen\altaffilmark{1},
T.~Aune\altaffilmark{2},
M.~Beilicke\altaffilmark{3},
W.~Benbow\altaffilmark{4},
A.~Bouvier\altaffilmark{2},
J.~H.~Buckley\altaffilmark{3},
V.~Bugaev\altaffilmark{3},
K.~Byrum\altaffilmark{5},
A.~Cannon\altaffilmark{6},
A.~Cesarini\altaffilmark{7},
L.~Ciupik\altaffilmark{8},
E.~Collins-Hughes\altaffilmark{6},
M.~P.~Connolly\altaffilmark{7},
W.~Cui\altaffilmark{9},
R.~Dickherber\altaffilmark{3},
J.~Dumm\altaffilmark{10},
A.~Falcone\altaffilmark{11},
S.~Federici\altaffilmark{12,13},
Q.~Feng\altaffilmark{9},
J.~P.~Finley\altaffilmark{9},
G.~Finnegan\altaffilmark{14},
L.~Fortson\altaffilmark{10},
A.~Furniss\altaffilmark{2},
N.~Galante\altaffilmark{4},
D.~Gall\altaffilmark{15},
S.~Godambe\altaffilmark{14},
S.~Griffin\altaffilmark{16},
J.~Grube\altaffilmark{8},
G.~Gyuk\altaffilmark{8},
J.~Holder\altaffilmark{17},
H.~Huan\altaffilmark{18},
G.~Hughes\altaffilmark{12},
T.~B.~Humensky\altaffilmark{19},
A.~Imran\altaffilmark{20},
P.~Kaaret\altaffilmark{15},
N.~Karlsson\altaffilmark{10},
M.~Kertzman\altaffilmark{21},
Y.~Khassen\altaffilmark{6},
D.~Kieda\altaffilmark{14},
H.~Krawczynski\altaffilmark{3},
F.~Krennrich\altaffilmark{20},
K.~Lee\altaffilmark{3},
A.~S~Madhavan\altaffilmark{20},
G.~Maier\altaffilmark{12},
P.~Majumdar\altaffilmark{1},
S.~McArthur\altaffilmark{3},
A.~McCann\altaffilmark{16},
P.~Moriarty\altaffilmark{22},
R.~Mukherjee\altaffilmark{23},
T.~Nelson\altaffilmark{10},
A.~O'Faol\'{a}in de Bhr\'{o}ithe\altaffilmark{6},
R.~A.~Ong\altaffilmark{1},
M.~Orr\altaffilmark{20},
A.~N.~Otte\altaffilmark{24},
N.~Park\altaffilmark{18},
J.~S.~Perkins\altaffilmark{25,26},
M.~Pohl\altaffilmark{13,12,31}
H.~Prokoph\altaffilmark{12},
J.~Quinn\altaffilmark{6},
K.~Ragan\altaffilmark{16},
L.~C.~Reyes\altaffilmark{27},
P.~T.~Reynolds\altaffilmark{28},
E.~Roache\altaffilmark{4},
J.~Ruppel\altaffilmark{13,12},
D.~B.~Saxon\altaffilmark{17},
M.~Schroedter\altaffilmark{4},
G.~H.~Sembroski\altaffilmark{9},
C.~Skole\altaffilmark{12},
A.~W.~Smith\altaffilmark{14},
I.~Telezhinsky\altaffilmark{13,12},
G.~Te\v{s}i\'{c}\altaffilmark{16},
M.~Theiling\altaffilmark{9},
S.~Thibadeau\altaffilmark{3},
K.~Tsurusaki\altaffilmark{15},
A.~Varlotta\altaffilmark{9},
M.~Vivier\altaffilmark{17},
S.~P.~Wakely\altaffilmark{18},
J.~E.~Ward\altaffilmark{3},
A.~Weinstein\altaffilmark{20},
R.~Welsing\altaffilmark{12},
D.~A.~Williams\altaffilmark{2},
B.~Zitzer\altaffilmark{5}
}

\and
\author{C. Pfrommer\altaffilmark{29,31},
A. Pinzke\altaffilmark{30}}

\altaffiltext{1}{Department of Physics and Astronomy, University of California, Los Angeles, CA 90095, USA}
\altaffiltext{2}{Santa Cruz Institute for Particle Physics and Department of Physics, University of California, Santa Cruz, CA 95064, USA}
\altaffiltext{3}{Department of Physics, Washington University, St. Louis, MO 63130, USA}
\altaffiltext{4}{Fred Lawrence Whipple Observatory, Harvard-Smithsonian Center for Astrophysics, Amado, AZ 85645, USA}
\altaffiltext{5}{Argonne National Laboratory, 9700 S. Cass Avenue, Argonne, IL 60439, USA}
\altaffiltext{6}{School of Physics, University College Dublin, Belfield, Dublin 4, Ireland}
\altaffiltext{7}{School of Physics, National University of Ireland Galway, University Road, Galway, Ireland}
\altaffiltext{8}{Astronomy Department, Adler Planetarium and Astronomy Museum, Chicago, IL 60605, USA}
\altaffiltext{9}{Department of Physics, Purdue University, West Lafayette, IN 47907, USA }
\altaffiltext{10}{School of Physics and Astronomy, University of Minnesota, Minneapolis, MN 55455, USA}
\altaffiltext{11}{Department of Astronomy and Astrophysics, 525 Davey Lab, Pennsylvania State University, University Park, PA 16802, USA}
\altaffiltext{12}{DESY, Platanenallee 6, 15738 Zeuthen, Germany}
\altaffiltext{13}{Institut für Physik und Astronomie, Universität Potsdam, 14476 Potsdam-Golm,Germany}
\altaffiltext{14}{Department of Physics and Astronomy, University of Utah, Salt Lake City, UT 84112, USA}
\altaffiltext{15}{Department of Physics and Astronomy, University of Iowa, Van Allen Hall, Iowa City, IA 52242, USA}
\altaffiltext{16}{Physics Department, McGill University, Montreal, QC H3A 2T8, Canada}
\altaffiltext{17}{Department of Physics and Astronomy and the Bartol Research Institute, University of Delaware, Newark, DE 19716, USA}
\altaffiltext{18}{Enrico Fermi Institute, University of Chicago, Chicago, IL 60637, USA}
\altaffiltext{19}{Physics Department, Columbia University, New York, NY 10027, USA}
\altaffiltext{20}{Department of Physics and Astronomy, Iowa State University, Ames, IA 50011, USA}
\altaffiltext{21}{Department of Physics and Astronomy, DePauw University, Greencastle, IN 46135-0037, USA}
\altaffiltext{22}{Department of Life and Physical Sciences, Galway-Mayo Institute of Technology, Dublin Road, Galway, Ireland}
\altaffiltext{23}{Department of Physics and Astronomy, Barnard College, Columbia University, NY 10027, USA}
\altaffiltext{24}{School of Physics \& Center for Relativistic Astrophysics, Georgia Institute of Technology, 837 State Street NW, Atlanta, GA 30332-0430}
\altaffiltext{25}{CRESST and Astroparticle Physics Laboratory NASA/GSFC, Greenbelt, MD 20771, USA.}
\altaffiltext{26}{University of Maryland, Baltimore County, 1000 Hilltop Circle, Baltimore, MD 21250, USA.}
\altaffiltext{27}{Physics Department, California Polytechnic State University, San Luis Obispo, CA 94307, USA}
\altaffiltext{28}{Department of Applied Physics and Instrumentation, Cork Institute of Technology, Bishopstown, Cork, Ireland}
% Christoph, Anders and Simone:
\altaffiltext{29}{Heidelberg Institute for Theoretical Studies, Schloss-Wolfsbrunnenweg 35, D-69118 Heidelberg, Germany}
\altaffiltext{30}{Department of Physics, University of California, Santa Barbara, CA 93106, USA}
\altaffiltext{31}{Corresponding authors: M. Pohl, pohlmadq@gmail.com \& C. Pfrommer, christoph.pfrommer@h-its.org}

@@

\begin{abstract}
Observations of radio halos and relics in galaxy clusters indicate efficient electron
acceleration. Protons should likewise be accelerated and, on account of
weak energy losses, can accumulate, suggesting that
clusters may also be sources of very high-energy (VHE; $E>100$ GeV) gamma-ray
emission. We
report here on VHE gamma-ray observations of the Coma galaxy cluster
with the VERITAS array of
imaging Cherenkov telescopes, with complementing \Fermi-LAT observations at
GeV energies. No significant gamma-ray emission from the Coma cluster was detected. Integral flux
upper limits at the 99\% confidence level were measured to be on the
order of $(2-5)\times 10^{-8}\ {\rm ph.\,m^{-2}\,s^{-1}}$ (VERITAS, $>220\ {\rm GeV}$) and $\sim 3\times 10^{-6}\ {\rm ph.\,m^{-2}\, s^{-1}}$
(\Fermi, $1-3\ {\rm GeV}$), respectively. We use the gamma-ray upper
limits to constrain CRs and magnetic fields in Coma. Using an analytical
approach, the CR-to-thermal pressure ratio is
constrained to be $< 16\%$ from VERITAS data and $< 2.5\%$ from \Fermi data (averaged within the
virial radius). Assuming that the radio-emitting electrons of the Coma halo result from hadronic
CR interactions, the observations imply a lower limit on the central magnetic field in Coma of
$1-3\,\mu{\rm G}$, depending on the radial magnetic-field profile and on the gamma-ray
spectral index. Since these values are well below those inferred by Faraday rotation measurements
in Coma, this makes the hadronic model a very plausible explanation of the Coma radio
halo. Finally, since galaxy clusters are dark-matter (DM) dominated, the VERITAS upper limits have
been used to place constraints on the thermally-averaged product of the total 
self-annihilation
cross section and the relative velocity of the DM particles, $\expval{\sigma v}$.
\end{abstract}

@@

The ICM is a hot ($T\sim 10^{8}$ K) plasma emitting thermal bremsstrahlung in the soft X-ray regime
\citep[see, e.g.,][]{article:Petrosian:2001}. This plasma has been heated primarily through
collisionless structure-formation shocks that form as a result of the hierarchical merging and
accretion processes. Such shocks and turbulence in the ICM gas in combination with intra-cluster
magnetic fields also provide a means to accelerate particles efficiently \citep[see,
e.g.,][]{article:ColafrancescoBlasi:1998, article:Ryu_etal:2003}. Many clusters feature megaparsec
scale halos of nonthermal radio emission, indicative of a population of relativistic electrons
and magnetic fields permeating the ICM \citep{article:Cassano_etal:2010}. There are two competing
theories to explain radio halos. In the ``hadronic model'', the radio-emitting
electrons and positrons are produced in inelastic collisions of cosmic-ray (CR) ions with the thermal gas of the
ICM \citep{article:Dennison:1980, article:EnsslinPfrommerMiniatiSubramanian:2011}. In the
``re-acceleration model'', a long-lived pool of 100-MeV electrons---previously accelerated by
formation shocks, galactic winds, or jets of active galactic nuclei (AGN)---interacts with plasma
waves that are excited during states of strong ICM turbulence, e.g., after a cluster merger. This
may result in second order Fermi acceleration and may produce energetic electrons ($\sim 10$ GeV)
sufficient to explain the observable radio emission \citep{article:SchlickeiserSieversThiemann:1987,
article:BrunettiLazarian:2010}. Observations of possibly nonthermal emission from clusters
in the extreme ultraviolet \citep[EUV; ][]{article:SarazinLieu:1998} and hard X-rays
\citep{article:RephaeliGruber:2002, article:Fusco-Femiano_etal:2004, article:Eckert_etal:2007} may
provide further indication of relativistic particle populations in clusters, although the
interpretation of these observations as nonthermal diffuse emission has been disputed on the basis
of more sensitive observations \citep[see, e.g.,][]{article:Ajello_etal:2009,
article:Ajello_etal:2010, article:Wik_etal:2009}.

@@

Gamma-ray emission can also be produced by Compton up-scattering of ambient photons, for example
CMB photons, on ultra-relativistic electrons. Those electrons can either be secondaries from the
CR interactions mentioned above,or injected into the ICM by powerful cluster members and
further accelerated by diffusive shock acceleration or turbulent reacceleration processes
\citep[][and references therein]{article:SchlickeiserSieversThiemann:1987}.

@@

\section{VERITAS Observations, Analysis, and Results}
The VERITAS gamma-ray detector \citep{article:Weekes_etal:2002} is an array of four 12 m-diameter
imaging atmospheric Cherenkov telescopes \citep{article:Holder_etal:2006} located at an
altitude of $\sim$1250 m a.s.l. at the Fred Lawrence Whipple Observatory in southern Arizona
(31$^{\circ}$~40\arcmin~30\arcsec~N, 110$^{\circ}$~57\arcmin~07\arcsec~W). Each of the telescopes
is equipped with a 499-pixel camera covering a 3.5$^{\circ}$ field of view. The array, completed in
the fall of 2007, is designed to detect gamma-ray emission from astrophysical objects in the
energy range from 100 GeV to more than 30 TeV. The energy resolution is $\sim 15$\% and the angular
resolution (68\% containment) is $\sim 0.1^{\circ}$ per event at 1 TeV. At the time of the Coma
cluster observations, the sensitivity of the array allowed for detection of a point source with a
flux of 1\% of the steady Crab Nebula flux above 300 GeV at the confidence level of five standard
deviations ($5\sigma$) in under 45 hours.\footnote{The integral flux sensitivity above 300 GeV was
improved by about 30\% with the relocation of one telescope in the summer of 2009.}

@@

No significant excess of VHE gamma rays from the Coma cluster was detected with VERITAS, as
illustrated by the $\theta^{2}$ distribution shown in Figure \ref{fig:thetasq}. The $\theta^{2}$
distribution is a radial comparison of the ON-source event distribution with the estimated
background. It extends out to 0.42 square degrees to cover both the case of point-like and extended
emission from the core of the Coma cluster. The $\theta^{2}$ distributions for the member galaxies
also considered in this work are very similar to that in Figure \ref{fig:thetasq} and show no
excess of gamma rays. A 99\% confidence level upper limit is calculated for each region of
interest using the method described by \citet{article:Rolke_etal:2005} assuming a
Gaussian-distributed background. A lower bound of zero is imposed on the gamma-ray flux from the Coma cluster, which prevents artificially low flux upper limits in the case that the best-fit source flux is formally negative. Figure \ref{fig:sigdist} shows the
distribution of significances over the VERITAS skymap, which is well fit by a Gaussian with a mean
close to zero and a standard deviation within a few percent of unity.

Table \ref{table:results} lists the upper limits for the selected regions of interest shown in
Table \ref{table:roi}. These upper-limit calculations depend on the gamma-ray spectrum, which in
this work is assumed to be a power law in energy, $dN/dE\propto E^{-\alpha}$, where the spectral index
$\alpha$ was allowed to have a value of 2.1, 2.3, or 2.5.

@@

\section{\Fermi-LAT Analysis and Results}
The Large Area Telescope (LAT) on board the Fermi Gamma-ray Space Telescope (Fermi) has observed
the Coma cluster in all-sky survey mode since its launch in June 2008. The Fermi-LAT is sensitive
to gamma rays in the 20 MeV to $\sim300$ GeV energy range and is very complementary to 
VERITAS. Ackermann et al. (2010) reported on the search for gamma-ray emission from
thirty-three galaxy clusters in the data from the first 18 months, including the Coma cluster, for
which an upper limit of $4.58\times10^{-9}$ ph. cm$^{-2}$ s$^{-1}$ in the 0.2 to 100 GeV energy band was reported. This limit is expected to improve as the exposure is increased. In this work an
updated analysis which includes additional data taken during the period 2010 February 5 to 2011 
June 10 is presented as a complement to the VERITAS results.

@@

The remaining background includes twenty-eight sources listed in the 11-month catalog of Fermi-LAT
data (Abdo et al. 2010), and the LAT standard Galactic diffuse emission component
\verb+gll_iem_v02.fit+ along with the corresponding isotropic template
\verb+isotropic_iem_v02.txt+ that accounts for extragalactic emission and residual instrumental
background. All of the twenty-eight sources are modeled as point sources with power-law energy
spectra. Fourteen of them lie outside the ROI and their spatial and spectral parameters were
kept fixed at the values given in the catalog. The parameters of the remaining sources, being
inside the ROI, and the normalization of the diffuse components were left to freely vary. The
gamma-ray emission from the Coma cluster is assumed to follow a power-law distribution with a
spectral index $\alpha$, which is allowed to have a value of 2.1, 2.3, or 2.5.

No significant gamma-ray signal was detected. Values of the significance greater than one were only
obtained for the high-energy bin, where the test statistic (TS) is $\sim2$ for the point source
model, $\sim4$ for the disk model with r$=0.2^{\circ}$ and $\sim6$ for the disk model with
r$=0.4^{\circ}$.

@@

\section{Gamma Ray Emission from Cosmic Rays}
We decided to adopt a multifaceted approach to constrain the CR-to-thermal pressure distribution in
the Coma cluster using the upper limits derived from the VERITAS and \Fermi-LAT data in this work.
This approach includes (1) a simplified multi-frequency analytical model that assumes a constant
CR-to-thermal energy density and a power-law spectrum in momentum, (2) an analytic model derived
from cosmological hydrodynamical simulations of the formation of galaxy clusters, and (3) a
model that uses the observed intensity profile
of the giant radio halo in Coma to place a lower limit on
the expected gamma-ray flux in the hadronic model -- where the radio-emitting electrons are
secondaries from CR interactions and which is independent of the magnetic field distribution.  This
last approach translates into a minimum CR pressure which, if challenged by tight gamma-ray
limits/detections, permits scrutiny of the hadronic interaction model of the formation of giant
radio halos. Alternatively, realizing a spatial CR distribution that is consistent with the flux
upper limits, and requiring the model to match the observed radio data, enables us to derive a
lower limit on the magnetic field distribution. We stress again that this approach assumes the
validity of the hadronic interaction model. Modeling the CR distribution through different
techniques enables us to bracket our lack of understanding about the underlying plasma physics that
shapes the CR distribution hence to reflect the Bayesian priors that are imposed on the modeling
\citep[see][for a discussion]{article:PinzkePfrommerBergstrom}.
